---
title: "Parallel Computing with R"
subtitle: Write a parallel for loop
week: 11
type: Case Study
---

```{r setup, include=FALSE, purl=F}
source("functions.R")
source("knitr_header.R")
library(knitr) 
```

```{r cache=F, message=F,warning=FALSE, results='hide'}
library(tidyverse)
library(spData)
library(sf)

## New Packages
library(mapview) # new package that makes easy leaflet maps
library(foreach)
library(doParallel)
registerDoParallel(4)
getDoParWorkers() # check registered cores
```

To use the tidycensus package, you will need to load the package and set your Census API key. A key can be obtained from http://api.census.gov/data/key_signup.html. You will only need to do that once (unless you delete your .Renviron file or move to a different computer).

```{r, eval=F}
# go to  http://api.census.gov/data/key_signup.html and get a key, then run the line below with your key.  Don't push your key to github!
library(tidycensus)
census_api_key("YOUR API KEY GOES HERE")
```

<div class="well">
<button data-toggle="collapse" class="btn btn-primary btn-sm round" data-target="#demo1">Show Hints</button>
<div id="demo1" class="collapse">

## Steps

Write an Rmd script that:

* Downloads block-level data on population by race in each census block in Buffalo using `get_dicennial()` function of the `tidycensus` package.  You can use the following code:
```{r, message=F}
library(tidycensus)
racevars <- c(White = "P005003", 
              Black = "P005004", 
              Asian = "P005006", 
              Hispanic = "P004003")

options(tigris_use_cache = TRUE)
erie <- get_decennial(geography = "block", variables = racevars, 
                  state = "NY", county = "Erie County", geometry = TRUE,
                  summary_var = "P001001", cache_table=T) 
```
* Crop the county-level data to `c(xmin=-78.9,xmax=-78.85,ymin=42.888,ymax=42.92)` to reduce the computational burdern. Feel free to enlarge this area if your computer is fast (or you are patient).
* Write a foreach loop that does the following steps _for each racial group_ in the `variable` column of the `erie` dataset and `rbind`s the results (e.g. `.combine=rbind`) into a single `sf` object.  You may want to convert the variable column into a factor and use `levels()` or use `unique()`.
   * filter the the data to include only one race at time
   * use `st_sample()` to generate random points for each person that resided within each polygon.  If you use a pipe (`%>%`), you will have to set `size=.$value`.  The `.` indicates that the column comes from the dataset that was passed to the function. See [here](https://magrittr.tidyverse.org/reference/pipe.html) for details on how to use the `.` in a pipe.
   * convert the points from `st_sample()` to spatial features with `st_as_sf()`
   * `mutate` to add a column named `variable` that is set to the current racial group (from the foreach loop)
* Use the `mapview()` function in the `mapview` package to make a leaflet map of the dataset and set the `zcol` to the racial identity of each point. You can adjust any of the visualization parameters (such as `cex` for size).  [Read more about mapview here](https://r-spatial.github.io/mapview/).  It's a new and really easy way to make leaflet maps from many types of spatial data.
   
</div>
</div>


```{r, purl=F, message=F, warning=F}
buffalo <- erie %>% 
   st_crop(c(xmin=-78.9,xmax=-78.85,ymin=42.888,ymax=42.92))



buffalo_dots=
   foreach(r=unique(buffalo$variable),.combine=rbind)%dopar%{
   filter(buffalo,variable==r) %>%  #do one variable at a time
   st_sample(size=.$value) %>%  #generate a point for each person randomly
   st_as_sf() %>%  #convert to a full sf object
   mutate(variable=r) #add the variable onto the points
   }
```


```{r eval=F, echo=F, purl=F, message=F, warning=F}
# Or, do it without pipes
buffalo_dots=
   foreach(r=unique(buffalo$variable),.combine=rbind)%dopar%{
   buffalo_race = filter(buffalo,variable==r)  #do one variable at a time
   buffalo_sample = st_sample(buffalo_race, 
                              size=buffalo_race$value)  #generate a point for each person randomly
   buffalo_sf = st_as_sf(buffalo_sample)  #convert to a full sf object
   mutate(buffalo_sf, variable=r) #add the variable onto the points
   }
```

Your final result should look something like this:
```{r, echo=F, purl=F, fig.width=2,fig.height=2, warning=F, as.is=T, cache=F}
mapview(buffalo_dots,zcol="variable",cex=1,lwd=0)
```

<div class="extraswell">
<button data-toggle="collapse" class="btn btn-link" data-target="#extras">
Extra time? Try this...
</button>
<div id="extras" class="collapse">

Update the map to include:

* Other racial groups
* Adjust colors to match the original
* Summarize the data in different ways (e.g. plot the polygon data, calculate indices, etc.)

```{r purl=F, echo=F}
```

</div>
</div>
